\documentclass[10pt]{amsart}
\usepackage[pdftex]{hyperref}
\usepackage{multicol}

\oddsidemargin=.15in
\evensidemargin=.15in
\textwidth=6in
\topmargin=-.4in
\textheight=9in
\parindent=0in
\pagestyle{empty}
%\input{tcilatex}

\begin{document}
 \thispagestyle{empty}

	\begin{center}
	{\LARGE {\bf Laura D'Angelo}} \\[.5pc]
	CURRICULUM VITAE
	\end{center}

	\bigskip 

	%{\large {\bf Personal Details}} \\*[-0.803pc]
	%\underline{\hspace{6in}} \newline
	%\newline
	%Date of Birth: December 25, 1981\newline
	%Place of Birth: Padova, Italy \newline
	%Nationality: Italian
	%\newline
	%\newline

	{\large {\bf Contact Information}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm] 
Department of Economics, Management and Statistics (DEMS)    \\
University of Milano-Bicocca     \\
Piazza dell'Ateneo Nuovo 1, \\
20126 Milan, Italy.           \\ \newline
e-mail: laura.dangelo@unimib.it
	\newline\newline

	{\large {\bf Current Position}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm] 
	{\it from December 2021}\newline
	{\bf Research fellow in Statistics
	} 
	\newline 
	Department of Economics, Management and Statistics (DEMS)    \\
	University of Milano-Bicocca
	\newline 

	{\large {\bf Research interests}} \\[-.08cm] 
	\underline{\hspace{6in}} 
	\begin{itemize}
	\item 	Statistical modeling
	\item 	Bayesian statistics
	\item	Bayesian nonparametrics
	\item 	Computational statistics
	\end{itemize}
		\vspace{2ex}
	{\large {\bf Education}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm] 
	{\it October 2018 - November 2021}\newline
	{\bf PhD in Statistical Science,  University of Padova. 
	} 
	\newline 
	{\it Thesis title: Bayesian modeling of calcium imaging data }\\
	Supervisor: Prof.\ Antonio Canale\\
	Co-supervisor: Prof.\ Michele Guindani.	\\ \newline
		{\it October 2015 -- November 2017 } \newline
	{\bf Master's degree ({\it laurea specialistica/magistrale}) in Statistical Science}.\newline
	University of Padova, Department of Statistics \newline
	Title of dissertation: ``Modelli Bayesiani nonparametrici: applicazioni al settore assicurativo   '' \newline
	 	Supervisor: Prof.\ Antonio Canale\\
	Final mark: 110/110 cum laude.
	\newline\newline
	 	{\it October 2012 -- September 2015} \newline
	{\bf Bachelor's degree ({\it laurea triennale}) in Statistics, Economics and Finance}. \newline
	University of Padova, Department of Statistics \\
	Title of dissertation: ``L’area sotto la curva ROC specifica per caratteristiche'' \newline
	Supervisor: Prof.\  Gianfranco Adimari\\
	Final mark:  110/110 cum laude.
	\newline\newline
	  
	{\large {\bf Visiting periods}} \\[-.08cm] 
	\underline{\hspace{6in}}\\[.2cm] 
	{\it January 2020 -- November 2020 }\newline
	University of California, Irvine;\\ Irvine, CA (USA).\\
	Supervisor: Prof.\ Michele Guindani. \newline
	\\



	 
	{\large {\bf Work experience}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm] 
	{\it January 2018 -- September 2018}\newline
	{\bf BIP S.p.A.}. \newline
	Consultant, data analyst.
	\newline\newline
	
	
 	{\large {\bf Awards and Scholarship}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm]
	{\it July 2021 - ISBA 2021 World Meeting.}\newline
	ISBA 2021 Best Student/Postdoc Contributed Paper Award.  
	\newline\newline
	
	
	 
	{\large {\bf Computer skills}} \\[-.08cm] 
	\underline{\hspace{6in}} 
	\begin{itemize}
	\item  R (advanced)
	\item  C++, GitHub, Python (working level)
	\end{itemize}
	
	$\,$\newline


	{\large {\bf Language skills}} \\[-.08cm] 
	\underline{\hspace{6in}} \newline
	Italian: native;\\
	English: fluent.  
	\newline
\newline
\newline
	 


%\clearpage
	{\large {\bf Publications}} \\[-.08cm] 
		\underline{\hspace{6in}} \\[.2cm] 
	{\bf Articles in journals}  \newline
	D'Angelo L. and Canale A. (2021) Contributed Discussion on: ``Centered partition processes: informative priors for clustering'', in \textit{Bayesian Analysis}, \textbf{16}(1), 356--358.
	%
	\newline\newline
	%
	D'Angelo L., Canale A., Yu Z. and Guindani M. (2021) Detection of neural activity in calcium imaging data via Bayesian mixture models, in \textit{Book of Short Papers SIS 2021} (Editors: Perna C., Salvati N., Schirripa Spagnolo F.), ISBN: 9788891927361.
	%
	\newline\newline
	%
	D'Angelo L. (2019) Model based clustering in group life insurance via Bayesian nonparametric mixtures, in \textit{Book of Short Papers SIS 2019} (Editors: Arbia, G., Peluso, S., Pini, A. and Rivellini, G.), ISBN: 978889191510.
	%
	\newline\newline
	%
	
%	{\bf Chapters in books}  \newline
%	Author1surname, Author1nameinitial, Author2surname, Author2nameinitial., \ldots    (year). Title (small initials).  In {\it Book Title} (capital initials), eds.\ Editor1initial., Editor1surname, Editor2initial., Editor2surname, \ldots, pp.\ firstpage--lastpage, Publisher, City.\newline\newline

	 
	{\bf Working papers}  \newline
	D'Angelo L., Canale A., Yu Z. and Guindani M. (2021) Bayesian nonparametric analysis for the detection of spikes in noisy calcium imaging data. \textit{arXiv preprint arXiv:2102.09403} 
	\newline\newline
	D’Angelo L. and Canale A. (2021) Efficient posterior sampling for Bayesian Poisson regression. \textit{arXiv preprint arXiv:2109.09520}
	\newline\newline
	
	
	{\large {\bf Conference presentations}} \\[-.08cm] 	\underline{\hspace{6in}}\\[.2cm] 
	D'Angelo L., Canale A., Yu Z. and Guindani M. (2021). Bayesian nonparametric analysis for the detection of spikes in noisy calcium imaging data. (contributed talk) \textit{JSM 2021}, August 8 - 12, 2021. 
	%
	\newline
	\newline
	%
	D'Angelo L., Canale A., Yu Z. and Guindani M. (2021). Bayesian nonparametric analysis for the detection of spikes in noisy calcium imaging data. (contributed talk) \textit{ISBA 2021 World Meeting}, June 23 - July 2, 2021. Pre-recorded video available at https://youtu.be/SLLSJVuFnMs .
	%
	\newline
	\newline
	%
	D'Angelo L., Canale A., Yu Z. and Guindani M. (2021). Detection of neural activity in calcium imaging data via Bayesian mixture models. (contributed talk) \textit{SIS 2021 Intermediate meeting}, Pisa, Italy, June 21 - 25, 2021.
	%
	\newline
	\newline
	%
	D'Angelo L. (2019). Model based clustering in group life insurance via Bayesian nonparametric mixtures. \textit{SIS 2019 Intermediate meeting}, Milan, Italy, June 12-14, 2019.
	\newline
	\newline
	\newline
	
	
	
	
	
	

	
	{\large {\bf Teaching experience}} \\[-.08cm] 
	\underline{\hspace{6in}} \\[.2cm] 
	{\it April 2021}\newline
	 Tirocinio formativo\\
%	 Degree\\
	 use of Latex for scientific writing, 2.5 hours\\
	 University of Padova
	 \newline\newline


%	{\it Month year -- Month year}\newline
%	 Course name\\
%	 Degree\\
%	 Teaching task (exercises/lab/\ldots), total number of hours\\
%	 Institution\\
%	 Instructor: Prof.\ Name Surname   \\
	  
	 
	 
%	 \ldots\ldots\\
%	  \ldots\ldots\\

 

	%\clearpage
	{\large {\bf Other Interests}} \\[-.08cm]%*[-0.806pc]
	\underline{\hspace{6in}} \\[.2cm] 
	Member of LIPU (Lega Italiana Protezione Uccelli) since 2019.\\
	Volunteer for the project Lipu LIFE Choo-na in 2019.
	\newline \newline

	%\newpage

	{\large {\bf References}} \\[-.08cm]%*[-0.806pc]
	\underline{\hspace{6in}} 
	\begin{multicols}{2}

	%\setlength{\columnsep}{50pt}
	%\setlength{\columnsep}{15cm}
	%\newline
	{\bf Prof. Antonio Canale}\newline
	University of Padova\\
	via Cesare Battisti, 241-243; \\
	35121 Padova. Italy.\\
	Phone: +39 049 827 4168\\
	e-mail: canale@stat.unipd.it  
	\newline\newline
	{\bf Prof. Michele Guindani}\newline
	University of California, Irvine\\
	Donald Bren School of Information and Computer Sciences;
	Irvine, CA 92697-1250\\
	Phone: +1 949 824 3276\\
	e-mail: michele.guindani@UCI.edu
	 	%\newline
	 		\end{multicols}




\end{document}
{"mode":"full","isActive":false}
